The Calculus of Linear Constructions

### 2021-10-26 writing/lics22/lics22.tex
lics22

  Linear Logic is a substructural logic introduced by Girard in his seminal work \cite{girard}. Girard notices that the weakening and contraction rules of Classical Logic when restricted carefully, give rise to a new logical foundation for reasoning about resource. Wadler \cite{wadler1990,wadler1991} first notices that an analogous restriction to variable usage in simple type theory leads to a linear type theory, where terms respect resources. A term calculus for linear type theory was later realized by Abramsky \cite{abramsky1993}. Benton \cite{benton1994} investigates the ramifications of the ! exponential in linear term calculi, decomposing it to adjoint connectives $F$ and $G$ that map between linear and non-linear judgments. Programming languages \cite{l3,ats,linear-haskell} featuring linear types have also been implemented, allowing programmers to write resource safe software in practical applications. The success of integrating Linear Logic with simple type theory exposes a tantalizing new frontier of integrating linearity with richer type theories.

  Work has been done to extend dependent type theories with linear types. Cervesato and Pfenning extend the Edinburgh Logical Framework with linear types \cite{lf,llf}, being the first to demonstrate that dependent types and linear types can coexist within a type theory. V\'{a}k\'{a}r \cite{vakar14} presents a linear dependent type theory, with syntax and semantics drawing inspiration from DILL \cite{dill}.  Krishnaswami et al. present a dependent linear type theory \cite{neel15} based on Benton's earlier work on mixed linear and non-linear calculus, demonstrating the ability to internalize imperative programming in the style of Hoare Type Theory \cite{htt}. Luo et al. introduce the property of essential linearity and a mixed linear/non-linear context, describing the first type theory that allows types to depend on linear terms. Based on initial ideas of McBride \cite{nothing}, Atkey's Quantitative Type Theory (QTT) \cite{qtt} uses semi-ring annotations to track variable occurrence, simulating irrelevance, linear and affine types within a unified framework. The Idris 2 programming language \cite{idris2} implements QTT as its core type system.

  We present a new linear dependent type system --- The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encode the usage of resources. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
  \begin{equation*}
    (x \ltype Money\ 5) \rightarrow (Camels + Marlboro)
  \end{equation*}
  This new specification for transaction demands a payment of 5 units of money, and the customer be relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

  Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``lightweight'' approach to extending CC$\omega$ with linear types, akin to Mazurak et al.'s work on System F \cite{mazurak}. This allows for a straightforward modeling of CLC in CC$\omega$, and lifting of CC$\omega$ into CLC, endowing CLC with the fruits of decades of CC research. We further extend CLC with inductive types, showing that CLC facilitates correct by construction imperative programming. We have formalized all major results in Coq, and implemented a prototype in OCaml.

  \medskip

  \noindent \textbf{\textit{Contributions}}: 
  Our contributions can be summarized as follows.
  \begin{itemize}
    \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
    \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, propagation and subject reduction.
    \item Next, we prove the linearity theorem, showing that the tangible impact linear types have how terms can be structured. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
    \item Additionally, we observe that CLC is highly backwards compatible with CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$. This shows that CLC is consistent.
    \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
    \item Furthermore, we extend CLC with linear inductive data, demonstrating CLC's capabilities for correct by construction imperative programming.
    \item Finally, we give an implementation extended with user definable linear and non-linear inductive types. Algorithmic type checking employed by the implementation streamlines the process of writing CLC.
  \end{itemize}

  \section{The Language of CLC}
  \subsection{Syntax}
  \begin{figure}[h]
    \vspace{-1.4em}
    \caption{Syntax}
    \vspace{1.4em}
    \begin{grammar}
      <$k$> := 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} predicative levels

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{4em} expressions
      \indalt $(x :_s A) \rightarrow B$
      \indalt $(x :_s A) \multimap B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \label{syntax}
  \end{figure}
  The syntax of the core type theory is presented in Figure \ref{syntax}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $k$ to specifically quantify over levels $0, 1, 2, ...$ that correspond to the predicative universes. We use the meta variable $i$ to quantify over all levels $*, 0, 1, 2, ...$. Here, $U_*$ is the impredicative universe of propositions, in the same spirit as CC$\omega$'s $Prop$ universe. $U_k$, and $L_k$ are the predicative universes of non-linear and linear types respectively.
  \begin{figure}[H]
    \vspace{-1.4em}
    \caption{Correspondence of CLC types and MELL implications}
    \Description{}
    \vspace{-1em}
    \begin{align}
      (\_ \utype A) \rightarrow B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ \ltype A) \rightarrow B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ \utype A) \multimap B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ \ltype A) \multimap B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \vspace{-1.6em}
    \label{correspondence}
  \end{figure}

@@

  \begin{figure}[H]
    \vspace{-1.6em}
    \caption{The Universe Hierarchy}
    \vspace{0.8em}
    \centering
    \begin{tikzcd}
      U_* \arrow[r, ":"', "\preceq", dash] 
      & U_0 \arrow[r, ":"', "\preceq", dash] 
      & U_1 \arrow[r, ":"', "\preceq", dash] 
      & U_2 \arrow[r, ":"', "\preceq", dash] 
      & \cdots \\
      & L_0 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_1 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_2 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & \cdots
    \end{tikzcd}
    \vspace{-0.5em}
    \label{universe}
  \end{figure}

@@

      \inferrule
      { \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        x \notin \Gamma_1 }
      { \mrg{\Gamma_1}
            {\Gamma_2, x \ltype A}
            {\Gamma, x \ltype A} }
      \rname{Merge-L2} 
    \end{mathpar}
    \vspace{-1em}
    \label{structural}
  \end{figure}

@@

  \begin{figure}[H]
    \vspace{-1em}
    \caption{Type Formation} 
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash U_* : U_0 } 
      \rname{Prop-Axiom}

@@

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A : L_k \\ 
        \Gamma \vdash B : s_k \\
        x \notin \Gamma }
      { \Gamma \vdash (x \ltype A) \multimap B : L_k } 
      \rname{L-Lolli}
    \end{mathpar}
    \vspace{-1em}
    \label{type}
  \end{figure}

@@

  \subsection{Term Formation} \label{teformation}
  The rules for term formation are presented in Figure \ref{term}.

  \begin{figure}[h]
    \caption{Term Formation}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma, x \utype A \vdash x : A } 
      \rname{U-Var}

@@

      \inferrule
      { \Gamma \vdash m : A \\
        \overline{\Gamma} \vdash B : s_i \\ A \preceq B }
      { \Gamma \vdash m : B } 
      \rname{Conv}
    \end{mathpar}
    \vspace{-1em}
    \label{term}
  \end{figure}

@@

      \inferrule
      { n \step n' }
      { m\ n \step m\ n' }
      \rname{Step-AppR}
    \end{mathpar}
    \label{parallel}
  \end{figure}
  As we have discussed previously, the elimination of the explicit ! exponential allows CLC to maintain the standard operational semantics of CC$\omega$, whose $\beta$-reductions are very well behaved.

  \subsection{Meta Theory} \label{meta}
  In this section, we focus our discussion on the meta-properties of CLC. First, we show the type soundness of CLC through the \textit{subject reduction} theorem. Next, we show that CLC is a valid linear type theory through the \textit{linearity} theorem. Furthermore, we show that the \textit{promotion} and \textit{dereliction} rules can be encoded as $\eta$-expansions. Finally, we construct a reduction preserving erasure function that maps well-typed CLC terms to well-typed CC$\omega$ terms, showing that CLC is strongly normalizing.

@@

  \subsubsection{Promotion and Dereliction}
  In most presentations of linear logic, non-linear objects are created by packing linear objects into the ! exponential using the \textit{promotion} rule. Conversely, the \textit{dereliction} rule unpacks non-linear objects so they can be used in place of linear ones. Linear type theories \cite{abramsky1993,dill,vakar14} utilizing exponential for non-linearity will contain ! types and rules that are equivalent to \textit{promotion} and \textit{dereliction}.

  Linear types of CLC are not obtained through packing and unpacking !, so there are no rules for \textit{promotion} and \textit{dereliction}. Arbitrary computations existing at the type level also muddle the association between which types can be promoted or derelicted to which. Nevertheless, \textit{promotion} and \textit{dereliction} for canonical types are derivable as theorems through $\eta$-expansion.

@@

  \begin{theorem}
    Lifting. If $\Gamma \vdash m : A$ is a valid judgment in CC$\omega$, then $\lift{\Gamma} \vdash \lift{m} \utype \lift{A}$ is a valid judgment in CLC.
  \end{theorem}

  \section{Extensions}
  Readers familiar with the literature of linear logic or linear type theory will have noticed at this point that CLC does not possess the famous $\otimes$ connective. Though it is entirely possible to encode $\otimes$ in core CLC through a linear Church encoding, proofs and programs written in this style will be incredibly difficult to maintain.

  A major design goal of CLC is to support an ergonomic interface for users to define inductive types \cite{inductive,cic} as an extension to the core theory. The presence of universe sorts $U$ and $L$ allows users to declare inductive types to be non-linear or linear. The lack of ! exponential simplifies the construction and elimination of inductive terms through direct use of constructors and pattern-matching. The $\otimes$ connective becomes definable as an inductive type.

@@

  Figure \ref{inductive} is an excerpt taken from our implementation's prelude, with syntax drawing heavily from Coq.

@@

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Formation Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \unit : U_0 }
      \rname{Unit-Form}

@@

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A : U_k \\ 
        \Gamma x \utype A \vdash B : L_k }
      { \Gamma \vdash F x:A.B : L_k }
      \rname{$F$-Form}
    \end{mathpar}
    \vspace{-0.4em}
    \label{formation}
    \Description{}
  \end{figure}

  \begin{figure}[h]
    \vspace{-0.8em}
    \caption{Select Introduction Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash () : \unit }
      \rname{Unit-Intro}

@@

      \inferrule
      { \pure{\Gamma_1} \\
        \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B[m/x] \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        \Gamma \vdash F x:A.B : L_k }
      { \Gamma \vdash \{m, n\} : F x:A.B }
      \rname{$F$-Intro}
    \end{mathpar}
    \vspace{-2em}
    \label{introduction}
    \Description{}
  \end{figure}

@@

  Though we have not implemented program extraction for CLC yet, our preliminary experiments show the feasibility of compiling the state axioms to constant time imperative operations after proof erasure. This can greatly improve the runtime efficiency of CLC over pure functional languages when used for programming. More complex examples are available in the first author's Github repository.

@@

  \begin{figure}[h]
  \vspace{-0.6em}
  \caption{Linear Lists}
  \begin{lstlisting}
  Inductive list (A : U) : L :=
  | nil : list A
  | cons : A -> list A -> list A.

  Fixpoint append (A : U) : list A -o 
    list A -o (list A -o list A) -o list A 
  :=
    fun ls1 ls2 k =>
      match ls1 with
      | nil => k ls2
      | (cons h t) as _cons_ => 
        append _ t ls2 
          (fun res => k (_cons_ h res))
      end.
  \end{lstlisting}
  \label{list}
  \vspace{-0.6em}
  \Description{}
  \end{figure}

  Figure \ref{list} presents a CPS append function for linear lists. In the \texttt{cons} case, the constructor is opened, exposing its data element \texttt{h} and tail list \texttt{t}. Instead of de-allocating the memory of \texttt{cons}, it is bound to the variable \texttt{_cons_} of linear type $A \multimap list\ A \multimap list\ A$. Basically, each \texttt{_cons_} must be applied exactly once to reconstruct a list using the memory of the original, thus mutating the original list in-place. Also notice that the continuation $k$ is of linear type $(list\ A \multimap list\ A)$, indicating that $k$ must be applied exactly once. Using linear types for continuations is surprisingly accurate as control-flow itself is linear.

@@

  \section{Conclusion}
  The Calculus of Linear Constructions is a lightweight extension of the Calculus of Constructions with linear types. As a type theory, it is a sound and effective framework for reasoning about resources. As a programing language, it allows one to write safe and efficient code, alleviating some of the common issues faced by standard functional programming.

>>>>

  This paper presents a new linear dependent type system --- The Calculus of Linear Constructions (CLC). CLC extends CC$\omega$ with linear types. CC$\omega$ itself is an extension of CC with a cumulative hierarchy of type universes. We add an extra universe sort $L$ of linear types with cumulativity parallel to the sort $U$ of non-linear types. This ultimately cumulates in the \textit{linearity} theorem, stating that all resources are used exactly once.

  The presence of both linear and dependent types enables CLC to write specifications that faithfully encode the usage of resources. The previous example of monetary transaction can be refined using an indexed linear type family $Money : \mathbb{N} \rightarrow L$ as follows.
  \begin{equation*}
    (x \ltype Money\ 5) \rightarrow (Camels + Marlboro)
  \end{equation*}
  This new specification for the transaction demands a payment of 5 units of money, and the customer be relieved of their ownership after the transaction finishes, effectively preventing the contradiction of having your cake and eating it too.

  Compared to preexisting approaches for integrating linear types and dependent types, CLC offers a ``lightweight'' approach to extending CC$\omega$ with linear types, akin to Mazurak et al.'s work on System F \cite{mazurak}. Prior works have either utilized a literal ! exponential or a pair of adjoint connectives to bridge the gap between linear and non-linear types, adding clutter that greatly hinders their practical usage. Our method requires neither exponential nor adjoint connectives, allowing for a straightforward modeling of CLC in CC$\omega$, and lifting of CC$\omega$ into CLC with minimal annotations, endowing CLC with the fruits of decades of CC research.

  We further extend CLC with inductive types, showing that CLC facilitates correct by construction imperative programming. We have formalized all major results in Coq, and implemented a prototype in OCaml.

  \noindent \textbf{\textit{Contributions}}: 
  Our contributions can be summarized as follows.
  \begin{itemize}
    \item Fist, we describe the Calculus of Linear Constructions, an extension to the Calculus of Constructions with linear types. The integration of linear types and dependent types allows CLC to directly and precisely reason about resource.
    \item Second, we study the meta-theory of CLC directly, showing that it satisfies the standard properties of confluence, propagation and subject reduction.
    \item Next, we prove the linearity theorem, showing the tangible impact linear types have on the structure of terms. We also show that promotion and dereliction of linear logic are derivable as theorems for canonical types.
    \item Additionally, we observe that CLC is highly backwards compatible with CC$\omega$, allowing us to construct a reduction preserving model of CLC in CC$\omega$, showing that CLC is consistent. We also prove that all valid CC$\omega$ terms after adding minimal amounts of annotation are valid CLC terms.
    \item All major results have been formalized and proven correct in the Coq Proof Assistant with help from the Autosubst \cite{autosubst} library. To the best of our knowledge, our development is the first machine checked formalization of a linear dependent type theory.
    \item Furthermore, we extend CLC with linear inductive data, demonstrating CLC's capabilities for correct by construction imperative programming.
    \item Finally, we give an implementation extended with user definable linear and non-linear inductive types. Algorithmic type checking employed by the implementation streamlines the process of writing CLC.
  \end{itemize}

  \section{The Language of CLC}
  \subsection{Syntax}
  \begin{figure}[h]
    \vspace{-1em}
    \caption{Syntax}
    \vspace{1em}
    \begin{grammar}
      <$k$> := 0 | 1 | 2 ... \phantom{* |} \hspace*{3em} predicative levels

@@

      <$m, n, A, B, C$> ::= $U_i$ | $L_k$ | $x$ \hspace*{4em} expressions
      \indalt $(x :_s A) \rightarrow B$
      \indalt $(x :_s A) \multimap B$
      \indalt $\lambda x. n$ | $m\ n$
    \end{grammar}
    \label{syntax}
  \end{figure}
  The syntax of the core type theory is presented in Figure \ref{syntax}. Our type theory contains two sorts of universes $U$ and $L$. We use the meta variable $k$ to specifically quantify over levels $0, 1, 2, ...$ that correspond to the predicative universes. We use the meta variable $i$ to quantify over all levels $*, 0, 1, 2, ...$. Here, $U_*$ is the impredicative universe of propositions, in the same spirit as CC$\omega$'s $Prop$ universe. $U_k$, and $L_k$ are the predicative universes of non-linear and linear types respectively.
  \begin{figure}[h]
    \vspace{-1em}
    \caption{Correspondence of CLC types and MELL implications}
    \Description{}
    \vspace{-1em}
    \begin{align}
      (\_ \utype A) \rightarrow B \quad &\Leftrightarrow \quad !(!A \multimap B) \\
      (\_ \ltype A) \rightarrow B \quad &\Leftrightarrow \quad !(A \multimap B) \\
      (\_ \utype A) \multimap B \quad &\Leftrightarrow \quad !A \multimap B \\
      (\_ \ltype A) \multimap B \quad &\Leftrightarrow \quad A \multimap B
    \end{align}
    \vspace{-1em}
    \label{correspondence}
  \end{figure}

@@

  \begin{figure}[H]
    \vspace{-1em}
    \caption{The Universe Hierarchy}
    \vspace{0.8em}
    \centering
    \begin{tikzcd}
      U_* \arrow[r, ":"', "\preceq", dash] 
      & U_0 \arrow[r, ":"', "\preceq", dash] 
      & U_1 \arrow[r, ":"', "\preceq", dash] 
      & U_2 \arrow[r, ":"', "\preceq", dash] 
      & \cdots \\
      & L_0 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_1 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & L_2 \arrow[r, "\preceq" description, dash] \arrow[ru, ":" description, dash] 
      & \cdots
    \end{tikzcd}
    \vspace{-0.5em}
    \label{universe}
    \Description{}
  \end{figure}

@@

      \inferrule
      { \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        x \notin \Gamma_1 }
      { \mrg{\Gamma_1}
            {\Gamma_2, x \ltype A}
            {\Gamma, x \ltype A} }
      \rname{Merge-L2} 
    \end{mathpar}
    \vspace{-1em}
    \label{structural}
    \Description{}
  \end{figure}

@@

  \begin{figure}[h]
    \caption{Type Formation} 
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash U_* : U_0 } 
      \rname{Prop-Axiom}

@@

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A : L_k \\ 
        \Gamma \vdash B : s_k \\
        x \notin \Gamma }
      { \Gamma \vdash (x \ltype A) \multimap B : L_k } 
      \rname{L-Lolli}
    \end{mathpar}
    \label{type}
    \Description{}
  \end{figure}

@@

  \subsection{Term Formation} \label{teformation}
  The rules for term formation are presented in Figure~\ref{term}.

  \begin{figure}[h]
    \caption{Term Formation}
    \begin{mathpar}
      \inferrule
      {\pure{\Gamma}}
      {\Gamma, x \utype A \vdash x : A} 
      \rname{U-Var}

@@

      \inferrule
      { \Gamma \vdash m : A \\
        \overline{\Gamma} \vdash B : s_i \\ A \preceq B }
      { \Gamma \vdash m : B } 
      \rname{Conv}
    \end{mathpar}
    \label{term}
    \Description{}
  \end{figure}

@@

      \inferrule
      { n \step n' }
      { m\ n \step m\ n' }
      \rname{Step-AppR}
    \end{mathpar}
    \label{parallel}
    \Description{}
  \end{figure}
  As we have discussed previously, the elimination of the explicit ! exponential allows CLC to maintain the standard operational semantics of CC$\omega$, whose $\beta$-reductions are very well behaved.

  \subsection{Meta Theory} \label{meta}
  In this section, we focus our discussion on the properties of CLC. First, we show the type soundness of CLC through the \textit{subject reduction} theorem. Next, we show that CLC is a valid linear type theory through the \textit{linearity} theorem. Furthermore, we show that the \textit{promotion} and \textit{dereliction} rules can be encoded as $\eta$-expansions. Finally, we construct a reduction preserving erasure function that maps well-typed CLC terms to well-typed CC$\omega$ terms, showing that CLC is strongly normalizing.

@@

  \subsubsection{Promotion and Dereliction}
  Linear types of CLC are not obtained through packing and unpacking !, so there are no explicit rules for \textit{promotion} and \textit{dereliction} of Linear Logic. Arbitrary computations existing at the type level also muddle the association between which types can be promoted or derelicted to which. Nevertheless, \textit{promotion} and \textit{dereliction} for canonical types are derivable as theorems through $\eta$-expansion.

@@

  \begin{theorem}
    Lifting. If $\Gamma \vdash m : A$ is a valid judgment in CC$\omega$, then $\lift{\Gamma} \vdash \lift{m} : \lift{A}$ is a valid judgment in CLC.
  \end{theorem}

  The lifting theorem shows that all valid CC$\omega$ terms are valid CLC terms if the domain of function types are annotated with sort $U$. In practice, even these annotations can be omitted if algorithmic type checking is able to infer them from context.

  \section{Extensions}
  Readers familiar with the literature of linear logic or linear type theory will have noticed at this point that CLC does not possess the iconic $\otimes$ connective. Though it is entirely possible to encode $\otimes$ in core CLC through a linear Church encoding, proofs and programs written in this style will be incredibly difficult to maintain.

@@

  A major design goal of CLC is to support an ergonomic interface for users to define inductive types \cite{inductive,cic} as an extension to the core theory. The presence of universe sorts $U$ and $L$ allow users to declare inductive types to be non-linear or linear. The lack of ! exponential simplifies the construction and elimination of inductive terms through direct use of constructors and pattern-matching. The $\otimes$ connective becomes definable as an inductive type.

  Figure \ref{inductive} is an excerpt taken from our implementation's prelude, with syntax heavily inspired by Coq.

@@

  \begin{figure}[H]
    \caption{Select Formation Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash \unit : U_0 }
      \rname{Unit-Form}

@@

      \inferrule
      { \pure{\Gamma} \\
        \Gamma \vdash A : U_k \\ 
        \Gamma x \utype A \vdash B : L_k }
      { \Gamma \vdash F x:A.B : L_k }
      \rname{$F$-Form}
    \end{mathpar}
    \label{formation}
    \Description{}
  \end{figure}

  \begin{figure}[H]
    \vspace{-0.8em}
    \caption{Select Introduction Rules}
    \begin{mathpar}
      \inferrule
      { \pure{\Gamma} }
      { \Gamma \vdash () : \unit }
      \rname{Unit-Intro}

@@

      \inferrule
      { \pure{\Gamma_1} \\
        \Gamma_1 \vdash m : A \\
        \Gamma_2 \vdash n : B[m/x] \\\\
        \mrg{\Gamma_1}{\Gamma_2}{\Gamma} \\
        \Gamma \vdash F x:A.B : L_k }
      { \Gamma \vdash \{m, n\} : F x:A.B }
      \rname{$F$-Intro}
    \end{mathpar}
    \vspace{-1em}
    \label{introduction}
    \Description{}
  \end{figure}

@@

  Though we have not implemented program extraction for CLC yet, our preliminary experiments show the feasibility of compiling the state axioms to constant time imperative operations after proof erasure. This can greatly improve the runtime efficiency of CLC over pure functional languages when used for programming. More complex examples such as safe array indexing are available in the first author's Github repository.

@@

  \begin{figure}[h]
  \caption{Linear Lists}
  \begin{lstlisting}
  Inductive list (A : U) : L :=
  | nil : list A
  | cons : A -> list A -> list A.

  Fixpoint append (A : U) : list A -o 
    list A -o (list A -o list A) -o list A 
  :=
    fun ls1 ls2 k =>
      match ls1 with
      | nil => k ls2
      | (cons h t) as _cons_ =>
        append _ t ls2 
          (fun res => k (_cons_ h res))
      end.
  \end{lstlisting}
  \label{list}
  \Description{}
  \end{figure}

  Figure \ref{list} presents a CPS append function for linear lists. In the \texttt{cons} case, the constructor is opened, exposing its data element \texttt{h} and tail list \texttt{t}. Instead of de-allocating the memory of \texttt{cons}, it is bound to the variable \texttt{_cons_} of linear type $A \multimap list\ A \multimap list\ A$. Basically, each \texttt{_cons_} must be applied exactly once to reconstruct a list using the memory of the original, thus mutating the original list in-place. Also notice that the continuation $k$ is of linear type $(list\ A \multimap list\ A)$, indicating that $k$ must be applied exactly once. Using linear types for continuations is surprisingly accurate as control-flow itself is a linear concept.

  \section{Related Work}
  Linear types are a class of type systems inspired by Girard' substructural Linear Logic \cite{girard}. Girard notices that the weakening and contraction rules of Classical Logic when restricted carefully, give rise to a new logical foundation for reasoning about resource. Wadler \cite{wadler1990,wadler1991} then applies an analogous restriction to variable usage in simple type theory, leading to the development of linear type theory where terms respect resources. A term calculus for linear type theory was later realized by Abramsky \cite{abramsky1993}. Benton \cite{benton1994} investigates the ramifications of the ! exponential in linear term calculi, decomposing it to adjoint connectives $F$ and $G$ that map between linear and non-linear judgments. Programming languages \cite{l3,ats,linear-haskell} featuring linear types have also been implemented, allowing programmers to write resource safe software in practical applications.

  Over the years, work has been done to enrich linear type theories with dependent types. Cervesato and Pfenning extend the Edinburgh Logical Framework with linear types \cite{lf,llf}, being the first to demonstrate that dependent types and linear types can coexist within a type theory. V\'{a}k\'{a}r \cite{vakar14} presents a linear dependent type theory, with syntax and semantics drawing inspiration from DILL \cite{dill}.  Krishnaswami et al. present a dependent linear type theory \cite{neel15} based on Benton's earlier work on mixed linear and non-linear calculus, demonstrating the ability to internalize imperative programming in the style of Hoare Type Theory \cite{htt}. Luo et al. \cite{luo} introduce the property of essential linearity and a mixed linear/non-linear context, describing the first type theory that allows types to depend on linear terms. Based on initial ideas of McBride \cite{nothing}, Atkey's Quantitative Type Theory (QTT) \cite{qtt} uses semi-ring annotations to track variable occurrence, simulating irrelevance, linear and affine types within a unified framework. The Idris 2 programming language \cite{idris2} implements QTT as its core type system.

@@
